---
title: "QSS Tutorial 3: Measurement I"
tutorial:
  id: "03-measurement1"
output:
  learnr::tutorial:
    progressive: true
    ace_theme: github
    theme: united
runtime: shiny_prerendered
description: >
  This tutorial will cover basic descriptive statistics and
  data visualization. Companion to Chapter 3 of "Quantitative Social
  Science: An Introduction."
---

```{r setup, include=FALSE}
library(gradethis)
library(learnr)
library(qsslearnr)
library(tidyverse)
tutorial_options(exercise.checker = gradethis::grade_learnr)
knitr::opts_chunk$set(echo = FALSE)
tut_reptitle <- "QSS Tutorial 3: Output Report"
data(STAR, package = "qss")
star <- STAR
```

## Handling Missing Data in R

### Small class size data


In this chapter, you'll analyze data from the STAR project, which is a four-year randomized trial on the effectiveness of small class sizes on education performance. The `star` data frame as been loaded into your space so that you can play around with it a bit.

#### Exercises
- Use the `slice` function on the `star` to see what the first 6 lines of the data look like.

```{r slice-star, exercise = TRUE}

```

```{r slice-star-solution}
star %>%
  slice(1:6)
```

```{r slice-star-check}
grade_code()
```

- Use the `dim` function on the `star` to see what the dimensions of the data look like.

```{r dim-star, exercise = TRUE}

```

```{r dim-star-solution}
dim(star)
```

```{r dim-star-check}
grade_code()
```


- Use the `summary` function on the `star` to get a sense for each variable.

```{r summary-star, exercise = TRUE}

```

```{r summary-star-solution}
summary(star)
```

```{r summary-star-check}
grade_code()
```



### Handling missing data


You probably noticed that there were some `NA` values in the data when you used the `slice()` function. These are **missing values**, where the value for that unit on that variable is missing or unknown. These values pose problems when we are trying to calculate quantities of interest like means or medians because R doesn't know how to handle them.

The first tool in your toolkit for missing data is the `is.na()` function. When you pass a vector `x` to `is.na(x)`, it will return a vector of the same length where each entry is `TRUE` if the value of `x` is `NA` and `FALSE` otherwise. Using logicals, you can easily get the opposite vector `!is.na(x)` which is `TRUE` when `x` is observed and `FALSE` when `x` is missing.

#### Exercises
- Use the `select`, `slice`, and `is.na` functions to show whether or not the first 6 values of the `g4math` variable from the `star` data frame are missing.

```{r headna, exercise = TRUE}

```

```{r headna-hint}
star %>%
  .. %>%
  .. %>%
  ..
```

```{r headna-solution}
star %>%
  select(g4math) %>%
  slice(1:6) %>%
  is.na()
```

```{r headna-check}
grade_result(
  pass_if(~ identical(.result,star %>%
  select(g4math) %>%
  slice(1:6) %>%
  is.na()))
)
```

- Use the `summarize`, `is.na`, and `sum` functions to show how many values of the `g4math` variable are missing.


```{r sumna, exercise = TRUE}

```


```{r sumna-hint-1}
star %>%
  summarize(missing = sum(..))

```

```{r sumna-hint-2}
star %>%
  summarize(missing = sum(is.na(..)))

```

```{r sumna-solution}
star %>%
  summarize(missing = sum(is.na(g4math)))

```

```{r sumna-check}
grade_code()
```


- Use the `summarize`, `is.na`, and `mean` functions to show what proportion of the `g4math` variable is missing.


```{r meanna, exercise = TRUE}

```

```{r meanna-hint-1}
star %>%
  summarize(missing = mean(..))
```

```{r meanna-hint-2}
star %>%
  summarize(missing = sum(is.na(..)))
```

```{r meanna-solution}
star %>%
  summarize(missing = sum(is.na(g4math)))
```

```{r meanna-check}
grade_code()
```



### Calculating means in the fact of missing data

Missing values makes it difficult to calculate numerical quantities of interest like the mean, median, standard deviation, or variance. Many of these function will simply return `NA` if there is a single missing value in the vector. We can instruct many function to ignore the missing values and do their calculation on just the observed data by using the `na.rm = TRUE` argument. For instance, suppose we have `x <- c(NA, 1,2,3)`, then `mean(x)` will return `NA`, but `mean(x, na.rm = TRUE)` will return `2`.

#### Exercises
- Try to calculate the `mean` of the `g4math` variable in the `star` data frame without setting `na.rm`.

```{r mean-with-na, exercise = TRUE}

```

```{r mean-with-na-solution}
mean(star$g4math)
```

```{r mean-with-na-check}
grade_code(
  correct = "This isn't that useful though!"
)
```

- Try to calculate the `mean` of the `g4math` variable when setting `na.rm = TRUE`.

```{r mean-narm, exercise = TRUE}

```

```{r mean-narm-solution}
mean(star$g4math, na.rm = TRUE)
```

```{r mean-narm-check}
grade_code()
```

## Visualizing Data

### Barplots

The **barplot** is a useful way to visualize a categorical or factor variable. In this exercise, you are going to visualize the `classtype` variable from the `star` data frame, which can take on the following values:

- `1` = small class
- `2` = regular class
- `3` = regular class with aid


#### Exercises
- Use `ggplot()` and `geom_bar()` to create a barplot of counts for each category of the `classtype` in the `star` data. Use `factor()` to turn `classtype` into a categorical variable.

```{r barplot, exercise = TRUE}
## tell R what data to use with ggplot, and...

# add the bar plot specify the x-axis classtype 

```

```{r barplot-hint-1}
## tell R what data to use with ggplot
ggplot(..) + 
# add the bar plot specify the x-axis classtype 
geom_bar(..)

```

```{r barplot-hint-2}
## tell R what data to use with ggplot
ggplot(data = ..) + 
# add the bar plot specify the x-axis classtype 
geom_bar(aes(x = factor(..)))

```

```{r barplot-solution}
## tell R what data to use with ggplot
ggplot(data = star) + 
# add the bar plot specify the x-axis classtype 
geom_bar(aes(x=factor(classtype)))
```

```{r barplot-check}
grade_code("Awesome. The graph is looking a little unhelpful, though. Let's spruce it up.")
```

### Making the barplot readable

The default barplot usually isn't all that readable.

#### Exercises

- Use `ggplot()` and `geom_bar()` as before, but now also...
- Use `scale_x_discrete()` with the `labels` input to change the x-axis labels to "Small class", "Regular class", and "Regular class with aid" (instead of 1, 2, 3).
- Use `xlab()` and `ylab()` to change the x-axis label to "Class type" and the y-axis label to "Number of students"

```{r barplot-labels, exercise = TRUE}
## tell R what data to use with ggplot, and...

# add the bar plot specify the x-axis classtype, and...

#  change the x-axis labels
  

# change the axis titles


```

```{r barplot-labels-hint-1}
## tell R what data to use with ggplot, and...
ggplot(data = star) + 
  
# add the bar plot specify the x-axis classtype, and...
geom_bar(aes(x=factor(classtype))) +
  
#  change the x-axis labels
  scale_x_discrete(..) +
  
# change the axis titles
  xlab(..) +
  ylab(..)
```

```{r barplot-labels-hint-2}
## tell R what data to use with ggplot, and...
ggplot(data = star) + 
  
# add the bar plot specify the x-axis classtype, and...
geom_bar(aes(x=factor(classtype))) +

#  change the x-axis labels
  scale_x_discrete(labels = c(..)) +

# change the axis titles
  xlab("Class type") +
  ylab(..)
```



```{r barplot-labels-solution}
## tell R what data to use with ggplot, and...
ggplot(data = star) + 
  
# add the bar plot specify the x-axis classtype, and...
geom_bar(aes(x=factor(classtype))) +

#  change the x-axis labels
  scale_x_discrete(labels = c("Small class", "Regular class", "Regular class with aid")) +
  
# change the axis titles
  xlab("Class type") +
  ylab("Number of students")
```

```{r barplot-labels-check}
grade_code("Great! So much better!")
```


### Histograms

For quantitative (numerical) variables, the barplot won't work because there are too many unique values. In this case, you will often use a histogram to visualize the a numerical variable.

#### Exercises
- Use the `geom_histogram` function to create a histogram for the `g4math` variable in the `star` data frame.
- Be sure to use the `...density...` as your `y-value` in the `aes()` function of `geom_histogram` to see the histogram with density

```{r hist, exercise = TRUE}
## create a histogram of g4math

```

```{r hist-hint}
## create a histogram of g4math
ggplot(star, aes(x = ..)) + 
  geom_histogram(aes(y = ..))
```


```{r hist-solution}
## create a histogram of g4math
ggplot(star, aes(x = g4math)) + 
  geom_histogram(aes(y = ..density..))
```

```{r hist-check}
grade_code("Great job, though the graph is a bit spartan. Let's make it more readable.")
```

### Sprucing up the histogram

There are several arguments you can add to improve or tweak our histogram:

Within `geom_histogram()`, we can add:

- `bins` or `binwidth`: a number specifying either how many bins our histogram should have or how wide each bin should be
- `boundary`: a number specifiying where a breakpoint should be (i.e., where one bin should end and another should begin)

We can also add more commands to our `ggplot()` equation:
- `ggtitle`: a character string that prints a main title for the plot.
- `xlim`, `ylim`: numeric vectors of length 2 to specify the interval for the x and y axes.

#### Exercises
- Create the same histogram as before, but now:
(a) set the `binwidth` to 5
(b) set the y-axis to be between `0` and `0.015` using the `ylim` argument, 
(c) include an informative x-axis label "Score" using the `xlab` argument, and 
(d) include a title ("Distribution of fourth-grade math scores") for the plot using the `ggtitle` argument.
- Make sure to separate the arguments in function calls with commas.


```{r hist-nice, exercise = TRUE}
## create the histogram with the specifications given in the instructions

```


```{r hist-nice-solution}
## create the histogram with the specifications given in the instructions
ggplot(star, aes(x = g4math)) + 
  geom_histogram(aes(y = ..density..), binwidth=5) +
  ylim(0, 0.015) +
  xlab("Score") +
  ggtitle("Distribution of fourth-grade math scores")

```

### Adding lines and text to a plot

We'll often want to add more information to a plot to make it even more readable. You can do that with commands that add to the current plot, such as `geom_vline` and `annotate`.
`geom_vline(xintercept = 1)` will add a vertical line to the plot at the specified value (`1` in this example).

#### Exercises
- Use the `geom_vline` function to add a vertical line at the mean of the `g4math` variable from the `star` data to our histogram from the last question. Remember, there are missing values in that variable so be sure to use the `na.rm = TRUE` argument to drop them.

```{r hist-line, exercise = TRUE}
ggplot(star, aes(x = g4math)) + 
  geom_histogram(aes(y = ..density..), binwidth=5) +
  ylim(0, 0.015) +
  xlab("Score") +
  ggtitle("Distribution of fourth-grade math scores") +

## add a vertical line at the mean of the variable

```

```{r hist-line-hint-1}
ggplot(star, aes(x = g4math)) + 
  geom_histogram(aes(y = ..density..), binwidth=5) +
  ylim(0, 0.015) +
  xlab("Score") +
  ggtitle("Distribution of fourth-grade math scores") +

## add a vertical line at the mean of the variable
  geom_vline(xintercept =..)

```

```{r hist-line-hint-2}
ggplot(star, aes(x = g4math)) + 
  geom_histogram(aes(y = ..density..), binwidth=5) +
  ylim(0, 0.015) +
  xlab("Score") +
  ggtitle("Distribution of fourth-grade math scores") +

## add a vertical line at the mean of the variable
  geom_vline(xintercept = mean(..))

```

```{r hist-line-solution}
ggplot(star, aes(x = g4math)) + 
  geom_histogram(aes(y = ..density..), binwidth=5) +
  ylim(0, 0.015) +
  xlab("Score") +
  ggtitle("Distribution of fourth-grade math scores") +

## add a vertical line at the mean of the variable
  geom_vline(xintercept = mean(star$g4math, na.rm=TRUE))


```

```{r hist-line-check}
grade_code()
```


### Adding text to a plot

We'll sometimes want to add text to a plot to make it more informative. `annotate(geom="text", label="z", x=x, y=y)` adds a character string `z` centered at point on the (`x`, `y`) on the plot. You can use the axis labels to see where you might want to add these parts of the plot.

#### Exercise

- Use the `annotate` function to add the string `Average Score` to the plot at the point (755, 0.0145).
- Make sure to separate the arguments in function calls with commas.

```{r hist-text, exercise = TRUE}
ggplot(star, aes(x = g4math)) + 
  geom_histogram(aes(y = ..density..), binwidth=5) +
  ylim(0, 0.015) +
  xlab("Score") +
  ggtitle("Distribution of fourth-grade math scores") + 

## add a vertical line at the mean of the variable
  geom_vline(xintercept = mean(star$g4math, na.rm=TRUE)) +

## add the text "Average Score" at the specified location

```

```{r hist-text-hint-1}
ggplot(star, aes(x = g4math)) + 
  geom_histogram(aes(y = ..density..), binwidth=5) +
  ylim(0, 0.015) +
  xlab("Score") +
  ggtitle("Distribution of fourth-grade math scores") + 

## add a vertical line at the mean of the variable
  geom_vline(xintercept = mean(star$g4math, na.rm=TRUE)) +

## add the text "Average Score" at the specified location
  annotate(..)
```

```{r hist-text-hint-2}
ggplot(star, aes(x = g4math)) + 
  geom_histogram(aes(y = ..density..), binwidth=5) +
  ylim(0, 0.015) +
  xlab("Score") +
  ggtitle("Distribution of fourth-grade math scores") + 

## add a vertical line at the mean of the variable
  geom_vline(xintercept = mean(star$g4math, na.rm=TRUE)) +

## add the text "Average Score" at the specified location
  annotate(geom=.., label=.., x=.., y=..)
```

```{r hist-text-solution}
ggplot(star, aes(x = g4math)) + 
  geom_histogram(aes(y = ..density..), binwidth=5) +
  ylim(0, 0.015) +
  xlab("Score") +
  ggtitle("Distribution of fourth-grade math scores") + 

## add a vertical line at the mean of the variable
  geom_vline(xintercept = mean(star$g4math, na.rm=TRUE)) +

## add the text "Average Score" at the specified location
  annotate(geom="text", label="Average Score", x=755, y=0.0145)
```

```{r hist-text-check}
grade_code()
```


## Submit

```{r context="setup"}
submission_ui
```

```{r context="server"}
submission_server()
```
