---
title: "QSS Tutorial 4: Measurement II"
tutorial:
  id: "04-measurement2"
output:
  learnr::tutorial:
    progressive: true
    ace_theme: github
    theme: united
runtime: shiny_prerendered
description: >
  This tutorial will cover basic descriptive statistics and
  data visualization.
---

```{r setup, include=FALSE}
library(gradethis)
library(learnr)
library(qsslearnr)
library(tidyverse)
tutorial_options(exercise.checker = gradethis::grade_learnr)
knitr::opts_chunk$set(echo = FALSE)
tut_reptitle <- "QSS Tutorial 4: Output Report"
data(STAR, package = "qss")
star <- STAR
```

## Conceptual Questions


```{r concept, echo=FALSE}
quiz(caption = "",
     question("In section 3.6 of QSS, what kind of plot are we using to visualize ideological differences between Democrats and Republicans?",
              answer("scatterplot", correct = TRUE),
              answer("bar plot"),
              answer("histogram")),
     question("To estimate the correlation between two variables x and y, we need to know the mean of each as well as the:",
              answer("maximum"),
              answer("standard deviation", correct = TRUE),
              answer("length")),
     question("What type of relationships does correlation measure?",
              answer("linear relationships", correct = TRUE),
              answer("nonlinear relationships"),
              answer("both"))
     )
```



## More visualization


We'll continue to analyze data from the STAR project, which is a four-year randomized trial on the effectiveness of small class sizes on education performance. The `star` data frame as been loaded into your space so that you can play around with it a bit.

### Boxplots

Boxplots are useful tools to visualize how the distribution of a continuous variable changes across levels of a categorical variable. 



#### Exercises

- Use `ggplot()` with the `star` dataframe (with the `x` aesthetic being a factor of `classtype` and the `y` aesthetic mapped to `g4math`) along with `geom_boxplot()` to make a boxplot.
- Include an informative x-axis label ("Class type"), an informative y-axis label ("Math scores"), and an informative title ("Distribution of Scores by Class Type")
- Use `scale_x_discrete` to set the labels for the x-axis ticks to "Small class", "Regular class", and "Regular class with aid"


```{r boxplot, exercise = TRUE}
## create a box plot with the characteristics specified in the instructions
ggplot(..) + 
  
```

```{r boxplot-hint-1}
ggplot(star, aes(..)) +
  geom_boxplot() + 
  xlab(..) + ylab(..) + ggtitle(..) +
  scale_x_discrete(..)
```

```{r boxplot-hint-2}
ggplot(star, aes(x = factor(..), y=..)) +
  geom_boxplot() + 
  xlab(..) + ylab(..) + ggtitle(..) +
  scale_x_discrete(labels=..)
```


```{r boxplot-hint-3}
ggplot(star, aes(x = factor(classtype), y=g4math)) +
  geom_boxplot() + 
  xlab("Class type") + ylab(..) + ggtitle(..) +
  scale_x_discrete(labels=c("Small class", .., ..))
```


```{r boxplot-solution}
ggplot(star, aes(x = factor(classtype), y=g4math)) +
  geom_boxplot() + 
  xlab("Class type") + ylab("Math scores") + ggtitle("Distribution of Scores by Class Type") +
  scale_x_discrete(labels=c("Small class", "Regular class", "Regular class with aid"))
```


```{r boxplot-check}
grade_code("Great!")
```


### Scatter plots

Now you'll think more about to measure bivariate relationships---that is, the relationship between two variables. The `geom_point(aes(x = x, y = y))` function will take two vectors and plot a series of points on a two-dimensional grid where the x-coordinates come from the `x` vector and the y-coordinates come from the `y` vector. You will use this plot to explore the relationship between math and reading test scores in the `star` data.

#### Exercises
- Create a scatter plot using the `plot` function with `g4math` on the x-axis and `g4reading` on the y-axis, both from the `star` data.
- Remember you can access a particular variable in a data frame using the `$` symbol.

```{r scatter, exercise = TRUE}
## produce a scatterplot of g4math on the x-axis and g4reading on the y-axis
ggplot(..) + geom_point(..)
```

```{r scatter-hint}
## produce a scatterplot of g4math on the x-axis and g4reading on the y-axis
ggplot(star) + geom_point(aes(x = .., y = ..))
```


```{r scatter-solution}
## produce a scatterplot of g4math on the x-axis and g4reading on the y-axis
ggplot(star) + geom_point(aes(x = g4math, y = g4reading))
```

```{r scatter-check}
grade_code("Ok, great plot. Let's make a bit more polished.")
```


### Plotting two sets of points

Often we want to plot certain pints differently than others. For instance, maybe you want to see how the relationship between math and reading scores differs between students in small classes versus those not in small classes. To do this, we can use subsets and the `points` function.

After using the `plot` function, you can add more points to the current plot by using the `points(x,y)` function, where the points are plotted very similarly to `plot`.

#### Exercises

- Create a subset of the `star` data called `small.class` which is when `classtype` is equal to 1. Create a second subset called `reg.class` which is when `classtype` is not equal to 1.
- Use the `plot` function to create a scatter plot of `g4math` and `g4reading` from the `small.class`. Change the color of the points by including the `col = "indianred"` argument.
- Use the `points` function to add the points of `g4math` and `g4reading` from the `reg.class`. Change the color of the points by including the `col = "dodgerblue"` argument.
- Remember that you can use `x == y` to test if `x` is equal to `y` and `x != y` to test if `x` is not equal to `y`

```{r plot-points, exercise = TRUE}
# take the data frame, and...
star %>% 
# create a new variable `class.size` that is "small" if classtype == 1 and "regular" if not, and...

  # pipe into ggplot, and...

# create a scatter plot with g4math on the x axis, g4reading on y, and class.size as color  


```

```{r plot-points-hint-1}
# take the data frame, and...
star %>% 
# create a new variable `class.size` that is "small" if classtype == 1 and "regular" if not, and...
  mutate(class.size = ..) %>%
# pipe into ggplot, and...
  ggplot() + 
# create a scatter plot with g4math on the x axis, g4reading on y, and class.size as color  
  geom_point(aes(x = .., y=.., color=..))
```

```{r plot-points-hint-2}
# take the data frame, and...
star %>% 
# create a new variable `class.size` that is "small" if classtype == 1 and "regular" if not, and...
  mutate(class.size = ifelse(classtype == 1, .., ..)) %>%
# pipe into ggplot, and...
  ggplot() + 
# create a scatter plot with g4math on the x axis, g4reading on y, and class.size as color  
  geom_point(aes(x = .., y=.., color=..))
```


```{r plot-points-hint-3}
# take the data frame, and...
star %>% 
# create a new variable `class.size` that is "small" if classtype == 1 and "regular" if not, and...
  mutate(class.size = ifelse(classtype == 1, "small", ..)) %>%
# pipe into ggplot, and...
  ggplot() + 
# create a scatter plot with g4math on the x axis, g4reading on y, and class.size as color  
  geom_point(aes(x = g4math, y=g4reading, color=..))
```


```{r plot-points-solution}
# take the data frame, and...
star %>% 
# create a new variable `class.size` that is "small" if classtype == 1 and "regular" if not, and...
  mutate(class.size = ifelse(classtype == 1, "small", "regular")) %>%
# pipe into ggplot, and...
  ggplot() + 
# create a scatter plot with g4math on the x axis, g4reading on y, and class.size as color  
  geom_point(aes(x = g4math, y=g4reading, color=class.size))
```


```{r plot-points-check}
grade_code("Looking good!")
```

### Finalizing your scatter plot

The scatter plot is looking very good, but it could use a little bit of polish. Let's add axis labels and a title. Remember that you can add these to the plot via the `xlab`, `ylab`, and `ggtitle`.

#### Exercises
- Take the current plot and add the following arguments: `xlab` ("Math scores"), `ylab` ("Reading scores"), and `ggtitle` ("Fourth Grade Math vs Reading Scores").

```{r niceplot, exercise = TRUE}
# take our prior plot...
star %>% 
  mutate(class.size = ifelse(classtype == 1, "small", "regular")) %>%
  ggplot() + 
  geom_point(aes(x = g4math, y=g4reading, color=class.size)) +
## add the arguments to make the plot more readable
```

```{r niceplot-hint}
# take our prior plot...
star %>% 
  mutate(class.size = ifelse(classtype == 1, "small", "regular")) %>%
  ggplot() + 
  geom_point(aes(x = g4math, y=g4reading, color=class.size)) +
## add the arguments to make the plot more readable
  xlab("Math scores") +
  ylab(..) +
  ggtitle(..)
```


```{r niceplot-solution}
# take our prior plot...
star %>% 
  mutate(class.size = ifelse(classtype == 1, "small", "regular")) %>%
  ggplot() + 
  geom_point(aes(x = g4math, y=g4reading, color=class.size)) +
## add the arguments to make the plot more readable
  xlab("Math scores") +
  ylab("Reading scores") +
  ggtitle("Fourth Grade Math vs Reading Scores")
```

```{r niceplot-check}
grade_code("Amazing!")
```


## Submit

```{r context="setup"}
submission_ui
```

```{r context="server"}
submission_server()
```
