---
title: "QSS Tutorial 2: Causality "
output:
  learnr::tutorial:
    progressive: true
    ace_theme: github
    theme: united
runtime: shiny_prerendered
description: >
  This tutorial will cover logicals, subsetting, factors, and
  calculating means in experiments. Companion to Chapter 2 of
  "Quantitative Social Science: An Introduction".
---


```{r setup, include=FALSE}
library(gradethis)
library(learnr)
library(qsslearnr)
tutorial_options(exercise.checker = gradethis::grade_learnr)
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
data(resume, package = "qss")
resume <- as_tibble(resume)
```

## Conceptual Questions


```{r concept-quiz, echo=FALSE}
quiz(
  caption = "",
  question(
    "Suppose a variable is binary, that is, it takes on values of either 0 or 1 (for example, female gender). Which of the following is the same as its sample mean?",
    answer("the sample median"),
    answer("the sample proportion of 1s", correct = TRUE),
    answer("neither of these")
  ),
  question(
    "What kind of value is `FALSE`?",
    answer("character"),
    answer("logical", correct = TRUE),
    answer("binary"),
    answer("numeric")
  ),
  question(
    "In order to calculate the mean of a variable we have used the `length()` function in the denominator. The `length()` of a vector is equivalent to:",
    answer("the number of elements", correct = TRUE),
    answer("the height"),
    answer("the maximum")
  ),
  question(
    "How are factor variables different from categorical variables?",
    answer("They are the same", correct = TRUE),
    answer("Factor variables contain numeric values"),
    answer("Categorical variables tend to have more levels or categories")
  )
)
```



## Exploring data

### Exploring the resume data

In this tutorial, we are going to be working with the resume data from Section 2.1 of QSS. This data comes from an experiment where researchers sent fictitious resumes with different names that implied different race and gender combinations to see if potential employers were more likely to call back names associated with different racial groups and genders.

Let's first explore the data a bit. It's stored as `resume`.

- Use the `head` function to show a preview of the `resume` data.


```{r head, exercise=TRUE}

```

```{r head-solution}
head(resume)
```

```{r head-check}
grade_code()
```

- Print out the dimension of the `resume` data using the `dim()` function


```{r resume-dim, exercise=TRUE}

```

```{r resume-dim-solution}
dim(resume)
```

```{r resume-dim-check}
grade_code()
```


- Use the `summary` function to show a summary of the data.


```{r res-summary, exercise=TRUE}

```

```{r res-summary-solution}
summary(resume)
```


```{r res-summary-check}
grade_code()
```



## Creating cross-tabs

### Counting units with unique combinations of values

To help you analyze this data, you can use a cross tabulation. Cross tabulation (or contingency table) is a table that quickly summarizes categorical data. For instance, in the resume data, we have a `sex` variable that tells us whether or not the fictitious resume had a male or a female name. 

First, we want to see how many units are in each combination of different variables. To do this in tidy, leverage `group_by` and `summarize`:

```{r, eval = FALSE, echo = TRUE}
mydata %>%
  group_by(var1, var2) %>%
  summarize(n = n())
```


Use this approach to count the number of values for each combination of the `sex` and `call` variables in the resume data.

```{r count, exercise = TRUE}

```

```{r count-hint}
resume %>%
  group_by(..., ...) %>%
  summarize(n = n())
```

```{r count-solution}
resume %>%
  group_by(sex, call) %>%
  summarize(n = n())
```


```{r count-check}
grade_code()
```

### Converting to cross-tab

The above output from `count` gave us four rows, one for each combination of the values of the two variables. What we often want, however, is to convert that to a 2 by 2 table called a cross tab. Cross-tabs are usually presented as tables where the rows correspond to values of one variable, columns to the values of another variable and the each cell gives the number of units for that combination of values. 

To do in a tidy way, we can reshape the output of the previous code using the `pivot_wider()` function:

```{r eval = FALSE, echo = TRUE}
mydata %>%
  group_by(var1, var2) %>%
  summarize(n = n()) %>%
  pivot_wider(names_from = var2, values_from = n)
```

The easiest way to use this function is specify a `names_from` argument which names the variable you want to move to the columns and a `values_from` argument that takes the name of variable where we should get the values to move. In our case for the a cross-tab, we'll want to move the `call` variable to the columns and get the values from the `n` column. 


Use this approach to turn the single-column of counts in the previous exercise into a 2x2 table with `sex` on the rows and `call` on the columns. 

```{r crosstab, exercise = TRUE}
resume %>%
  group_by(sex, call) %>%
  summarize(n = n())
```

```{r crosstab-hint}
resume %>%
  group_by(sex, call) %>%
  summarize(n = n()) %>%
  pivot_wider(names_from = ..., values_from = ...)
```

```{r crosstab-solution}
resume %>%
  group_by(sex, call) %>%
  summarize(n = n()) %>%
  pivot_wider(names_from = call, values_from = n)
```

```{r crosstab-check}
grade_code()
```

### Proportions in crosstabs

Another task we often want to do is to convert the counts in a cross-tab into proportions of the overall data. To do that, all we have to do is modify the `summarize()` call using `nrow()`:


```{r eval = FALSE, echo = TRUE}
mydata %>%
  group_by(var1, var2) %>%
  summarize(n = n() / nrow(mydata)) %>%
  pivot_wider(names_from = var2, values_from = n)
```

Take the code from the previous cross-tab and turn it into a table of proportions. 

```{r crosstab_prop, exercise = TRUE}
resume %>%
  group_by(sex, call) %>%
  summarize(n = n()) %>%
  pivot_wider(names_from = call, values_from = n)
```

```{r crosstab_prop-hint}
resume %>%
  group_by(..., ...) %>%
  summarize(n = n() / ...) %>%
  pivot_wider(names_from = ..., values_from = ...)
```

```{r crosstab_prop-solution}
resume %>%
  group_by(sex, call) %>%
  summarize(n = n() / nrow(resume)) %>%
  pivot_wider(names_from = call, values_from = n)
```

```{r crosstab_prop-check}
grade_code()
```


## Means and Differences in Means by Groups

### Calculating a mean by groups 

We often need to calculate the means of a variable (usually the outcome) within levels of another variable. To do this, we can use the same `group_by()` and `summarize()` approach:

```{r eval = FALSE, echo = TRUE}
mydata %>%
  group_by(group_var) %>%
  summarize(across(outcome_var, mean))
```

Here, `group_var` is the grouping variable and `outcome_var` is the variable we want to take the average of. 

Calculate the average of the `call` variable within levels of the `race` variable.

```{r avg_by_group, exercise = TRUE}

```

```{r avg_by_group-hint}
resume %>%
  group_by(...) %>%
  summarize(across(..., mean))
```

```{r avg_by_group-solution}
resume %>%
  group_by(race) %>%
  summarize(across(call, mean))
```

```{r avg_by_group-check}
grade_code()
```

### Pulling the means into a vector

Now, once we have the means, we may want to take differences between them. To do this, it's often easiest to use the `pull()` function after `summarize()` to get a vector with just the means:

```{r, eval = FALSE, echo = TRUE}
mydata %>%
  group_by(group_var) %>%
  summarize(across(outcome_var, mean)) %>%
  pull(outcome_var, name = group_var)
```

The first argument to `pull()` is the column name of the variable you want to pull out (the outcome variable in this case). The second argument, `names` is the name of the variable that should provide the names of the vector. 

Use this approach to create a vector (not a tibble) of the means of `call` within levels of `race`:

```{r avgs_vec, exercise = TRUE}
resume %>%
  group_by(race) %>%
  summarize(across(call, mean))
```

```{r avgs_vec-hint}
resume %>%
  group_by(race) %>%
  summarize(across(call, mean)) %>%
  pull(..., name = ...)
```

```{r avgs_vec-solution}
resume %>%
  group_by(race) %>%
  summarize(across(call, mean)) %>%
  pull(call, name = race)
```

```{r avgs_vec-check}
grade_code()
```

### Calculating the difference in means/average treatment effect

A common task in experiments or any data analysis is to compute the difference in means between two or more groups. We can do this with the output of the last step, where we have a vector of group means that are named with the values of the groups. From there, we can take differences between elements in the vector. For instance, let's say `group_var` has levels `a` and `b` and we want to know the difference in means between levels `a` and `b`. 

```{r, eval = FALSE, echo = TRUE}
avgs <- mydata %>%
  group_by(group_var) %>%
  summarize(across(outcome_var, mean)) %>%
  pull(outcome_var, name = group_var)

avgs["b"] - avgs["a"]
```

Now, use this approach to calculate the difference in mean callbacks (`call`) by race (`race`). Remember that the values of `race` is `"black"` and `"white"`.

```{r diff_in_mean, exercise = TRUE}

```

```{r diff_in_mean-hint}
avgs <- resume %>%
  group_by(race) %>%
  summarize(across(call, mean)) %>%
  pull(call, name = race)

avgs[...] - avgs[...]
```

```{r diff_in_mean-solution}
avgs <- resume %>%
  group_by(race) %>%
  summarize(across(call, mean)) %>%
  pull(call, name = race)

avgs["black"] - avgs["white"]
```

```{r diff_in_mean-check}
grade_code()
```

### Calculating difference in means within groups

Another common task is to calculate the difference in means between groups *within levels of another group*. For instance, we might want to know the effect of treatment within levels of gender or ideology. This is slightly more tricky than the last approach since we can't just use `pull()`. Instead, we'll need to use `pivot_wider` after we summarize and then create a new variable that calculates the differences. 

```{r, eval = FALSE, echo = TRUE}
mydata %>%
  group_by(var1, var2) %>%
  summarize(across(outcome_var, mean)) %>%
  pivot_wider(names_from = var1, values_from = outcome_var) %>%
  mutate(diff_in_means = b - a)
```

Again, here we have assumed that the levels of `var1` are `a` and `b`.

Now, let's apply this to the resume data and find the difference in callback means between Black and White resumes within levels of the gender of the resume. Remember that the values of `race` is `"black"` and `"white"`.

```{r diff_by_gender, exercise = TRUE}

```

```{r diff_by_gender-hint}
resume %>%
  group_by(race, ...) %>%
  summarize(across(call, mean)) %>%
  pivot_wider(names_from = ..., values_from = ...) %>%
  mutate(diff_in_means = ... - ...)
```


```{r diff_by_gender-solution}
resume %>%
  group_by(race, sex) %>%
  summarize(across(call, mean)) %>%
  pivot_wider(names_from = race, values_from = call) %>%
  mutate(diff_in_means = black - white)
```

```{r diff_by_gender-check}
grade_code()
```

## More Logicals in R

### Complex relationals

In this exercise, you have the ages of a sample of 15 people, stored in the `ages` vector. We can use these relational operators to create a logical vector which indicates which ages fall within a specific range. In particular, we can find out which respondents are college-aged (18-22).

#### Exercise

- Create a logical vector, called `college.aged`, which indicates which observations in `ages` are greater than or equal to 18 and less than or equal to 22. Be sure to use parentheses to separate out the two logical statements.
- Take the sum of the `college.aged` vector to determine how many 18-22 year olds there are in the sample.

```{r ages, exercise = TRUE, exercise.setup = "generate-ages"}
## check the value of the ages vector
ages <- c(31, 20, 43, 45, 41, 46, 28, 49, 61, 19, 39)

## create a logical vector called college.aged
## that is TRUE for someone between 18-22, inclusive


## find the number of college.aged respondents

```

```{r ages-solution}
college.aged <- (ages >= 18) & (ages <= 22)
sum(college.aged)
```

```{r ages-check}
grade_result_strict(
  pass_if(~ identical(college.aged, (ages >= 18) & (ages <= 22))),
  pass_if(~ identical(.result, sum((ages >= 18) & (ages <= 22))))
)
```


### Subsetting based on logicals


In the last exercise, you used logical statements to create a vector that told us whether each entry in the `ages` vector is in the 18-22 year-old range. We can now use that information to figure out what the actual ages of the respondents in that range are.

#### Exercise

- Use the brackets and the `college.aged` logical vector that has already been created to subset `ages` to the value only between 18 and 22, inclusive.
- Use the `mean` function to calculate the average age of the respondents in this subset. To do this, use the bracket subsetting and the `college.aged` logical vector that has already been created.

```{r generate-ages}
ages <- c(31, 20, 43, 45, 41, 46, 28, 49, 61, 19, 39)
```

```{r logsubset, exercise = TRUE, exercise.setup = "generate-ages"}
## here's the college.aged logical vector
college.aged <- (ages >= 18) & (ages <= 22)

## calculate the average age among the college-aged in the sample

```

```{r logsubset-solution}
mean(ages[college.aged])
```

```{r logsubset-check}
grade_result(
  pass_if(~ identical(.result, mean(ages[college.aged])))
)
```

#### Exercise

Try to do the same thing with tidyverse notation. `ages` has been turned into a data frame for you with one column `ages` to work properly with tidyverse functions.

-  Use `filter()` to subset `ages` to the value only between 18 and 22, inclusive.
- Use the `summarize()` function to calculate the average age of the respondents in this subset.

```{r generate-ages-tidy}
ages <- c(31, 20, 43, 45, 41, 46, 28, 49, 61, 19, 39)
ages <- data.frame(ages)
```

```{r logsubset-tidy, exercise = TRUE, exercise.setup = "generate-ages"}
## calculate the average age among the college-aged in the sample

```

```{r logsubset-tidy-hint}
ages %>%
  filter(.., ..) %>%
  summarize(mean = ..) %>% 
  pull()
```


```{r logsubset-tidy-solution}
ages %>%
  filter(ages >= 18, ages <= 22) %>%
  summarize(mean = mean(ages)) %>% 
  pull()
```

```{r logsubset-tidy-check}
grade_code()
```


### Using simple conditional statements

For this exercise, we'll use the resume data once again.

What if we wanted to create a new vector that depends on whether a statement is true or false? For example, suppose you wanted to create an indicator variable for whether or not a specific resume had the name "Carrie." We can obtain this using the `ifelse(X, Y, Z)` command inside of `mutate()`. This command takes a logical vector as `X` and returns a new vector of the same length as `X` that has the value `Y` if that value in `X` is TRUE and `Z` if that value in `X` is FALSE.

#### Exercise

- Use the `ifelse` function to create a new variable called `carrie` that is 1 if the resume name (`firstname`) is `"Carrie"` and 0 otherwise.
- Print the first six lines of `resume` using the `head` function to see the new variable.

```{r ifelse, exercise = TRUE}
## create a new variable called carrie
resume <- resume %>%
  mutate(carrie = ..)

## print the first 6 lines of the updated resume

```

```{r ifelse-hint}
## create a new variable called carrie
resume <- resume %>%
  mutate(carrie = ifelse(.., 1, 0))
## print the first 6 lines of the updated resume
head(resume)
```



```{r ifelse-solution}
## create a new variable called carrie
resume <- resume %>%
  mutate(carrie = ifelse(firstname == "Carrie", 1, 0))
## print the first 6 lines of the updated resume
head(resume)
```

```{r ifelse-check}
grade_code()
```

## Factors variables in R

### Factor variables


You have seen that creating subsets can be helpful for calculating different quantities or statistics for specific subgroups in the data. When there is more than 1 or 2 subgroups of interest, however, this can be a cumbersome process. For that reason, it's helpful to know about factor variables. Basically, a factor variable is a categorical variable that takes a finite number of distinct values.

Any variable can be turned into a factor by calling the `as.factor()` function like so:

    mydata$myvar <- as.factor(mydata$myvar)

This will take the variable `myvar` and create a factor variable with levels that are observed in that variable. Most often, you will convert a character variable to a factor.

#### Exercise

- Finish the code below that creates the `type` character variable. Fill in the last values of `race` and `sex` and add the label `WhiteMale` to this last type.
- Convert the `resume$type` variable to a factor variable using `as.factor()`.

```{r factor, exercise = TRUE}
## fill in the last line of code to create a character vector for the type of
## application that was sent
resume_fact <- resume %>%
  mutate(type = if_else(race == "black" & sex == "female", "BlackFemale", ""),
         type = if_else(race == "black" & sex == "male", "BlackMale", type),
         type = if_else(race == "white" & sex == "female", "WhiteFemale", type),
         type = if_else(race == .. & sex == .., .., type)) %>%
  mutate(type = ..) #turn into a factor
```

```{r factor-hint}
## fill in the last line of code to create a character vector for the type of
## application that was sent
resume_fact <- resume %>%
  mutate(type = if_else(race == "black" & sex == "female", "BlackFemale", ""),
         type = if_else(race == "black" & sex == "male", "BlackMale", type),
         type = if_else(race == "white" & sex == "female", "WhiteFemale", type),
         type = if_else(race == .. & sex == .., "WhiteMale", type)) %>%
  mutate(type = as.factor(..)) #turn into a factor
```

```{r factor-solution}
## fill in the last line of code to create a character vector for the type of
## application that was sent
resume_fact <- resume %>%
  mutate(type = if_else(race == "black" & sex == "female", "BlackFemale", ""),
         type = if_else(race == "black" & sex == "male", "BlackMale", type),
         type = if_else(race == "white" & sex == "female", "WhiteFemale", type),
         type = if_else(race == .. & sex == .., "WhiteMale", type)) %>%
  mutate(type = as.factor(type)) #turn into a factor

```

```{r factor-check}
grade_code("Fantastic, you got that factor loaded up and ready to go. Now, let's see what you can do with it.")
```


In tidyverse, we can actually create this factor more quickly using `case_when()`, which allows us to supply different assignments when different conditional statements are met. As an example, let's say we have a new vector of ages, ranging from 14 to 30. We want to assign whether we think these people are likely in high school (ages 15-17), college (18-21), or neither (22+). We can use `case_when()` to fill out our options. The `~` separates our conditional from what we want the column to be populated with. We end with `TRUE ~ "neither"` as a catchall for all other observations.

```{r}
ages <- data.frame(age = c(15, 18, 22, 16, 20, 29, 21, 27))

ages <- ages %>% 
  mutate(type = case_when(age>=15 & age < 18 ~ "highschool",
                          age >=18 & age <22 ~ "college",
                          TRUE ~ "neither"))

```


#### Exercise

- Using `case_when()`, repeat the previous exercise: create a the `type` character variable for each gender and race combination and then turn `type` into a factor.

```{r casewhen, exercise = TRUE}
resume_cw <- resume %>%
mutate(type = case_when(race == .. & sex == .. ~ ..,
                        race ==  .. & sex == .. ~ ..,
                        race == .. & sex == .. ~ ..,
                        race == .. & sex == .. ~ ..,
                        .. ~ "other")) %>%
  mutate(type = ..) #turn into a factor
```

```{r casewhen-hint}
resume_cw <- resume %>%
mutate(type = case_when(race == "black" & sex == "female" ~ "BlackFemale",
                        race ==  .. & sex == .. ~ ..,
                        race == .. & sex == .. ~ ..,
                        race == .. & sex == .. ~ ..,
                        TRUE ~ "other")) %>%
  mutate(type = as.factor(..)) #turn into a factor
```

```{r casewhen-solution}
## fill in the last line of code to create a character vector for the type of
## application that was sent
resume_cw <- resume %>%
mutate(type = case_when(race == "black" & sex == "female" ~ "BlackFemale",
                        race == "white" & sex == "female" ~ "WhiteFemale",
                        race == "black" & sex == "male" ~ "BlackMale",
                        race == "white" & sex == "male" ~ "WhiteMale",
                        TRUE ~ "other")) %>%
  mutate(type = as.factor(type)) #turn into a factor
```

```{r casewhen-check}
grade_code("Great!!!")
```




## Submit

```{r context="setup"}
submission_ui
```

```{r context="server"}
submission_server()
```
